MSstatsPTM statistical relative quantification of post-translational modifications in bottom-up proteomic experiments

### 2022-03-21 main/MSstatsPTM_Main.tex
Background

Mass spectrometry (MS)-based proteomics is increasingly used to detect changes in post-translational modifications (PTMs) between samples from various conditions. Analysis of data from such experiments faces numerous statistical challenges. These include the low abundance of modified proteoforms, the small number of representative peptides that span modification sites, and confounding between changes in the abundance of PTM and the overall changes in the protein abundance. Therefore, statistical approaches for detecting differential PTM abundance must integrate all the available information pertaining to a PTM site, and consider all the relevant sources of confounding and variation. In this manuscript we propose such a statistical framework, which is comprehensive, accurate, and leads to reproducible results. The framework requires an experimental design, which for each sample quantifies both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites. The framework supports both label-free and tandem mass tag (TMT)-based acquisitions. The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design. Next, model-based inferences regarding the PTM and the protein-level abundances are combined to account for the confounding between these two sources. Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments demonstrate the improved fold change estimation and detection of differential PTM abundance, as compared to currently used approaches. The proposed framework is implemented in the free and open-source R/Bioconductor package $MSstatsPTM$.

@@

This manuscript proposes a general statistical analysis framework that detects relative changes in post-translational modifications. The framework requires an experimental design, which for each sample quantifies both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites. The framework supports both label-free (DIA, DDA) and tandem mass tag (DDA-TMT)-based acquisitions. The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design. Next, model-based inferences regarding the PTM and the protein-level abundances are combined to account for the confounding between these two sources.

We evaluated the proposed framework on two datasets from computer simulations, one benchmark controlled mixture, and three biological investigations. The datasets illustrate a diverse set of organisms, acquisition methods, and experimental designs, showing the applicability of the framework to a variety of situations. The evaluations demonstrated that by appropriately leveraging the information from the unmodified peptide, the proposed approach improves the accuracy of the estimates of PTM fold changes and results in a better calibrated false positive rate of detecting differentially abundant PTMs. In particular, accounting for the confounding from protein abundance allows us to characterize the true effect of the modification, avoiding the need for more manual and time intensive methods.

@@

The datasets in this manuscript  benchmarked the proposed approach in situations with known ground truth, and represented a variety of experimental designs and acquisition methods. Two computer simulations varied in experimental realism. The first simulation produced a perfectly clean dataset, with many replicates and no missing values. The second simulation introduced real-world characteristics, such as limited modified features and missing values. The spike-in experiment took the real world characteristics a step further and allowed us to compare the methods in a real experiment with known changes in modified spike-in peptides. Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, PTM types, experimental designs and acquisition strategies. Table \ref{tab:dataDescription} summarizes the experiments. Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and results of the statistical analysis are available in MassIVE.quant (\url{https://massive.ucsd.edu/ProteoSAFe/static/massive-quant.jsp}) \cite{Choi:2020}. Details of computer simulations are available on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).

@@

The simulation provides an idealistic case. 24 synthetic datasets were generated exhibiting scenarios with different experimental designs and biological variation. 1000 proteins were generated per dataset with 10 unmodified features per protein. One PTM was created for each protein in the dataset, resulting in 1000 PTMs, each with 10 modified features. 500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions. The fold change of half the 500 differential PTMs were fully masked by changes in the unmodified protein. Additionally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified protein. All differential PTMs were generated with an expected log fold change of .75 between conditions. Each simulation was generated with random biological variation. Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$. The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$. The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$. The values for $\sigma^2$ were motivated by the biological experiments in this manuscript.
Further details can be found in Supplementary Sec. \ref{supp-sec:dataset1}.

@@

The data were simulated as above, while including missing values and a low number of modified features to provide a more realistic representation of the experiments. In this simulation, the feature counts and the proportion of missing values were as observed on average over all the the experimental datasets in this manuscript. PTMs were simulated with 2 modified peptide features, while unmodified proteins were simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random. \todo{There will likely be a comment that missing values are more common among low-abundance features. Did the simulation account for that? Devon: The simulation did not include any missing values due to low abundance. Should we mention this or just leave it as is?} Further details can be found in Supplementary Sec. \ref{supp-sec:dataset2}.

@@

In the custom spike-in experiment 8 biological replicates were allocated to 4 conditions, resulting in 2 biological replicates per condition, and a balanced design. Additional information can be found in Supplementary Sec. \ref{supp-sec:benchmark}.

\medskip \noindent {\bf Experimental Design:} 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark.  Unmodified peptides from Human lysate were used to estimate changes in global protein abundance. Background {\it E. coli} lysate was used to normalize total protein levels prior to enrichment or global protein profiling. The background lysate was treated as control, i.e. it had no expected changes in any comparison.

@@

\subsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
\label{sec:exp_proc_dataset4}
In this experiment 11 biological replicates were allocated to 1 TMT mixture in an unbalanced design. Further details are in Supplementary Sec. \ref{supp-sec:ipah}.

\medskip \noindent {\bf Experimental Design:} Luchetti et al. \cite{LUCHETTI2021} quantified the abundance of total protein and ubiquitination in human epithelial cells. Cells were engineered to express IpaH7.8 under a dox inducible promoter and measurements were taken at different time periods. GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The proteome was quantified using isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS. \todo{One MS run total? Only one MS run was used. Not sure the best way to say this.}

@@

\medskip \noindent {\bf Pairwise comparisons:} We evaluated the ability of the statistical methods to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance. The six condition were labeled Dox1hr, Dox2hr, Dox4hr, Dox6hr, NoDox0hr, and NoDox6hr. All conditions were compared with each other, resulting in 15 pairwise comparisons. Since the dataset was a biological investigation, the true positive modifications were unknown.

\subsection*{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
\label{sec:exp_proc_dataset5}
In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced design. More details are in Supplementary Sec. \ref{supp-sec:shigella}.

@@

The proteome was quantified in 11-plex isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS. The six conditions were split between 11 channels, and as the result the experiment was unbalanced. Each mixture contained two replicates per early and late WT and KO conditions. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT.

@@

In this experiment 8 biological replicates were split between 4 conditions in a balanced design. Further details in Suppler Sec \ref{supp-sec:usp30}.

@@

\medskip \noindent {\bf Global Profiling Run:} This experiment did not include a separate global profiling run to measure unmodified peptides. In this case $MSstatsPTM$ can still be used by extracting unmodified peptides from the modified run. However, in addition to low feature counts for unmodified peptides,  this leads to substantially fewer matches between modified and unmodified peptides. There was a 41.9\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Background Information}

\subsection*{Experimental Design}

\todo{Would it be possible to re-organize this section? (1) Overview, including the data structures  the figure and the notation - essentially the section that describes Fig 1; (2) Existing methods: limma and $t$-test, including the description in the notation of the figure; also Isobar-PTM - can express in the same notation? Ok if not; (3) Existing methods: MSstats. Not applicable for PTM but useful background; (4) Proposed approach in the notation of the figure. I feel that the experimental design in Suppler is redundant. Could we fold it into the main text as well?} \todo{In this way, the entire Suppler Sec 2 will be merged into main \; the only methods in supplementary will be the table }

\todo{Devon: I moved most everything around as recommended. The notation is mainly described once we start getting into the equations (t-test section). We can move this into the design section but I thought it was clearer to describe it with the equations. The results section now starts with the proposed method in the same notation. The supplementary now only includes the table and extra results.}

Figure \ref{fig:data-structure} schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present. A PTM site is quantified with multiple spectral features, which vary in sequence (e.g., fully or partially cleaved peptides), ionization efficiency, charge states, etc. Additionally, some spectral features can be missing. The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition $i$ is denoted by $i^{\ast}$.  The number of quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. To perform adjustment with respect to protein abundance, features of unmodified peptides are used for the inference of underlying protein abundance. Features corresponding to unmodified peptides are considered together, where the protein abundance in Condition $i$ is denoted by $i$. Typically, because of the enrichment step for PTMs, very few of those features are present in original LC-MS runs. For more accurate estimation of protein abundance, separate global proteomics data of unenriched samples are recommended. If unmodified peptides are unavailable for any given modification, unmodified intensity adjustment cannot be performed.

@@

Despite the importance of PTMs for biological function, there is a lack of general statistical analysis framework that summarizes the available quantitative information from LC-MS/MS experiments, performs statistical inference, and characterizes the quantitative properties of PTM in a statistically rigorous manner.

Many investigations performed differential abundance analysis of PTMs using two-sample $t$-test or its extensions.  The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}. Modifications of the $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}. While simple, these approaches do not fully account for the sources of variations, and are not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.

Isobar-PTM was also proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}. Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level. Unfortunately, this statistical modeling framework is not applicable to either label-free workflows or experiments with complex designs.

An overview of the existing approach using $t$-test and Limma are shown in the following sections.

\subsubsection*{Two-sample $t$-test}

Two-sample $t$-test is based on the null hypothesis that there is no difference in mean PTM abundance between Conditions $i$ and $i^*$. There are $I$ conditions and $J$ mass spectrometry runs (technical replicates) per condition in the experiment. The PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$. The abundance in each run is taken as input and is often estimated by sum of peak intensities. For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by
\[
\log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right).
\]
For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate, 
and the $t$-test for the adjusted PTM abundance on log scale takes as input the difference of their log-estimates.
%which is equivalent to take the difference of their log abundances. 
The quantity is denoted by $d_{ij}$ and is given by
\[
d_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right).
\]
Alternatively, run-level summarization can also be used. The difference between the means of PTM abundance in Conditions $i$ and $i'$ are estimated as
\[
\hat{\Delta} = \frac{1}{J} d_{i+}^{\ast} - \frac{1}{J} d_{i'+},
\]
where $d_{i+}^{\ast} = \sum_{j=1}^{J} d_{ij}^{\ast}$ and the test statistic for the $t$-test is given by $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$. The statistical significance of the difference is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.

\subsubsection*{Limma}

@@

\subsection*{Existing work in statistical modeling and parameter estimation}

The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}. $MSstats$ takes as input a list of log-transformed intensities of spectral features which are used to characterize the identified protein. For each protein, the feature log-intensities are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. The model parameters are estimated using a split-plot approach, where the feature log-intensities are first summarized into a single value per protein per run in the subplot model, and the protein-level summaries are then used for the inference of the protein abundance. This approach allows the method to be extended to cases where the experimental design is unbalanced and where additional sources of variation are present. In the protein-level summarization, Tukey's median polish (TMP), a simple and robust procedure, is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each protein~\cite{Tukey:1977}. After summarization, the inference of the protein abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}. We leverage this statistical modeling and quantification workflow and apply it to PTMs, in addition to unmodified proteins.

@@

@@

All datasets for the four experiments and two simulations, including raw files and analysis results, are available online.

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.6]{images/fig3.png}
\caption{
Data structure of a typical PTM experiment and goals of PTM characterization. (a) Schematic data representation, in a simplified case of two conditions and two replicate runs. Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide. The feature intensities the PTM in Condition i are denoted by i*. Features corresponding to unmodified peptides in Condition i is denoted by i. Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines). (b) PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty. (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
\label{fig:data-structure}
\end{figure}

>>>>

Liquid chromatography coupled with mass spectrometry (LC-MS/MS)-based proteomics is increasingly used to detect changes in post-translational modifications (PTMs) between samples from various conditions. Analysis of data from such experiments faces numerous statistical challenges. These include the low abundance of modified proteoforms, the small number of representative peptides that span modification sites, and confounding between changes in the abundance of PTM and the overall changes in the protein abundance. Therefore, statistical approaches for detecting differential PTM abundance must integrate all the available information pertaining to a PTM site, and consider all the relevant sources of confounding and variation. In this manuscript we propose such a statistical framework, which is versatile, accurate, and leads to reproducible results. The framework requires an experimental design, which for each sample quantifies both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites. The framework supports both label-free and tandem mass tag (TMT)-based LC-MS/MS acquisitions. The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design. Next, model-based inferences regarding the PTM and the protein-level abundances are combined to account for the confounding between these two sources. Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments demonstrate the improved fold change estimation and detection of differential PTM abundance, as compared to currently used approaches. The proposed framework is implemented in the free and open-source R/Bioconductor package $MSstatsPTM$.

@@

This manuscript proposes a general statistical analysis framework that detects relative changes in post-translational modifications. The framework requires an experimental design, which for each sample quantifies both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites. The framework supports data-dependent acquisitions (DDA) that are label-free or tandem mass tag (TMT)-based. The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design. Next, model-based inferences regarding the PTM and the protein-level abundances are combined to account for the confounding between these two sources.

We evaluated the proposed framework on two datasets from computer simulations, one benchmark controlled mixture, and three biological investigations. The datasets illustrate a diverse set of organisms, acquisition methods, and experimental designs, showing the applicability of the framework to a variety of situations. The evaluations demonstrated that by appropriately leveraging the information from the unmodified peptides, the proposed approach improves the accuracy of the estimates of PTM fold changes and results in a better calibrated false positive rate of detecting differentially abundant PTMs. In particular, accounting for the confounding from protein abundance allows us to characterize the true effect of the modification, avoiding the need for more manual and time intensive methods.

@@

The datasets in this manuscript  benchmarked the proposed approach in situations with known ground truth, and represented a variety of experimental designs and acquisition methods. Two computer simulations varied in experimental realism. The first simulation produced a perfectly clean dataset, with many replicates and no missing values. The second simulation introduced real-world characteristics, such as limited modified features and missing values. The spike-in experiment took the real world characteristics a step further and allowed us to compare the methods in a real-world experimental settings with known changes in modified spike-in peptides. Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, PTM types, experimental designs and acquisition strategies. Table \ref{tab:dataDescription} summarizes the experiments. Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and results of the statistical analysis are available in MassIVE.quant (\url{https://massive.ucsd.edu/ProteoSAFe/static/massive-quant.jsp}) \cite{Choi:2020}. Details of computer simulations are available on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).

@@

The simulation represented an idealistic case. 24 synthetic datasets were generated exhibiting scenarios with different experimental designs and biological variation. In each dataset, 1000 proteins had 10 unmodified features per protein. Each of the 1000 proteins had one PTM, each PTM represented by 10 modified features. The PTMs of 500 proteins had a differential fold change between conditions, while the other 500 proteins were generated with no difference between conditions. Furthermore, the fold changes of half of the 500 differential PTMs were fully masked by changes in the unmodified protein. Additionally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified protein. All differential PTMs were generated with an expected log fold change of 0.75 between conditions.

Each simulation was generated with random biological variation. Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$. The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$. The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$. The values for $\sigma^2$ were motivated \todo{motivated how specifically? corresponded to median variances?}  by the biological experiments in this manuscript.
Further details are in Supplementary Sec. \ref{supp-sec:dataset1}.

@@

The data were simulated as above, while including missing values and a low number of modified features to provide a more realistic representation of the experiments. In this simulation, the feature counts and the proportion of missing values were as observed on average over all the the experimental datasets in this manuscript. Specifically, PTMs were simulated with 2 modified peptide features, while unmodified proteins were simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random. \todo{There will likely be a comment that missing values are more common among low-abundance features. Did the simulation account for that? Devon: The simulation did not include any missing values due to low abundance. Should we mention this or just leave it as is?} Further details are in Supplementary Sec. \ref{supp-sec:dataset2}.

@@

In the custom spike-in experiment 8 biological \todo{I am not sure these are bio reps. If this is the same lysate, then these are in fact technical reps (there is no between-subject variation)} replicates were allocated to 4 conditions, resulting in 2 biological replicates per condition, and a balanced design. Additional information can be found in Supplementary Sec. \ref{supp-sec:benchmark}.

\medskip \noindent {\bf Experimental Design:} 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark.  Unmodified peptides from Human lysate were used to estimate changes in global protein abundance. Background {\it E. coli} lysate was used to normalize total protein levels prior to enrichment or global protein profiling. The background lysate was treated as control, i.e. it had no expected changes in any comparison. \todo{I am a bit confused. Fig S7 shows that there is a difference in Background between Mix 1,2 and Mix 3,4. Could you explain better? In fact, would it be possible to move the figure here for clarity, and have the figure as panel (a), and the table as panel (b)? And in the text above refer to the figure. This will help the understanding} \todo{Also, why is it a balanced design - as stated in the paragraph above - when the background is not present everywhere?}

@@

\subsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
\label{sec:exp_proc_dataset4}
In this experiment 11 biological replicates were allocated to 1 TMT mixture in an \todo{group comparison or time course?} unbalanced design. Further details are in Supplementary Sec. \ref{supp-sec:ipah}.

\medskip \noindent {\bf Experimental Design:} Luchetti et al. \cite{LUCHETTI2021} quantified the abundance of total protein and ubiquitination in human epithelial cells. Cells were engineered to express IpaH7.8 under a dox inducible promoter and measurements were taken at different time periods. GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. \todo{If this is a group comparison, explain why, even though the conditions are times. The table above says it is group comparison} The proteome was quantified using isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS. \todo{One MS run total? Only one MS run was used. Not sure the best way to say this.}

@@

\medskip \noindent {\bf Pairwise Comparisons:} We evaluated the ability of the statistical methods to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance. The six condition were labeled Dox1hr, Dox2hr, Dox4hr, Dox6hr, NoDox0hr, and NoDox6hr. All conditions were compared with each other, resulting in 15 pairwise comparisons. Since the dataset was a biological investigation, the true positive modifications were unknown.

\subsection*{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
\label{sec:exp_proc_dataset5}
In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced \todo{time course} design. More details are in Supplementary Sec. \ref{supp-sec:shigella}.

@@

The proteome was quantified in 11-plex isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS. The six conditions were split between 11 channels, and as the result the experiment was unbalanced. Each mixture contained two replicates per early and late WT and KO conditions. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT. \todo{was there a reference channel?}

@@

In this experiment 8 biological replicates were split between 4 conditions in a balanced \todo{group comparison} design. Further details in Suppler Sec \ref{supp-sec:usp30}.

@@

\medskip \noindent {\bf Global Profiling Run:} This experiment did not include a separate global profiling run to measure unmodified peptides. In this case $MSstatsPTM$ could still be used by extracting unmodified peptides from the modified run. However, in addition to low feature counts for unmodified peptides,  this lead to substantially fewer matches between modified and unmodified peptides. There was a 41.9\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Background}

\subsection*{Goals of PTM characterization, input to statistical analyses, and notation}

Consider an LC-MS/MS experiment with $I$ conditions and $J$ mass spectrometry runs (technical \todo{of biological?} replicates) per condition in the experiment. For one protein, the PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$.

Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one protein in a simplified experiment with $I=2$ conditions. \todo{Describe the population parameters in the figure (quantities of interest}.  The data are collected in $J=2$ \todo{biological?} replicate runs \todo{describe how the data are representative of the population parameters of interest}
The number of quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. \todo{There may be outliers}. Some spectral features can be missing.

\todo{in Figure 2 b, there are two quantities of interest, stated at the population level. Describe the quantities, what is of interest, why relevant}

%---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

@@

\noindent {\bf Two-sample $t$-test} \todo{When describing the existing methods, could you refer to Fig 1? Here is the suggested order: (1) What is the null hypothesis? Do these methods test the same null hypothesis? (If not, it may be a good idea to add this null hypothesis to the figure for contrast) (2) What are the data (how summarized prior to the testing); (3) what is the model assumptions? (4) how inference is performed} Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions.

\todo{These sentences were combined from different places. Could you edit?} The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}. Two-sample $t$-test is based on the null hypothesis that there is no difference in mean PTM abundance between Conditions $i$ and $i^*$. The abundance in each run is taken as input and is often estimated by sum of peak intensities. For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by
\[
\log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right).
\]
For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate, 
and the $t$-test for the adjusted PTM abundance on log scale takes as input the difference of their log-estimates.
%which is equivalent to take the difference of their log abundances. 
The quantity is denoted by $d_{ij}$ and is given by
\[
d_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right).
\]
Alternatively, run-level summarization can also be used. The difference between the means of PTM abundance in Conditions $i$ and $i'$ are estimated as
\[
\hat{\Delta} = \frac{1}{J} d_{i+}^{\ast} - \frac{1}{J} d_{i'+},
\]
where $d_{i+}^{\ast} = \sum_{j=1}^{J} d_{ij}^{\ast}$ and the test statistic for the $t$-test is given by $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$. The statistical significance of the difference is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.

\medskip \noindent {\bf Limma} \todo{Could you fine-tune, using the same order as above for Limma? I.e., null hypothesis, the data, the model, the inference} Modifications of the $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}. While simple, these approaches do not fully account for the sources of variations, and are not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.

@@

\medskip \noindent {\bf Isobar-PTM} was also proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}. Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level. Unfortunately, this statistical modeling framework is not applicable to either label-free workflows or experiments with complex designs.

\subsection*{Statistical modeling and parameter estimation in MSstats}

\todo{Could you also refer to Figure 1? Essentially, MSstats only analyzes a part of the figure. Refer to the null hypothesis. Then describe the summarization (including AFT etc), and the model for he summarized data (the fact that different designs require different models - this is your argument for the limitation of the methods above.)} The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}. $MSstats$ takes as input a list of log-transformed intensities of spectral features which are used to characterize the identified protein. For each protein, the feature log-intensities are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. The model parameters are estimated using a split-plot approach, where the feature log-intensities are first summarized into a single value per protein per run in the subplot model, and the protein-level summaries are then used for the inference of the protein abundance. This approach allows the method to be extended to cases where the experimental design is unbalanced and where additional sources of variation are present. In the protein-level summarization, Tukey's median polish (TMP), a simple and robust procedure, is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each protein~\cite{Tukey:1977}. After summarization, the inference of the protein abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}. We leverage this statistical modeling and quantification workflow and apply it to PTMs, in addition to unmodified proteins.

@@

\todo{Although demonstrated here on DDA, is also applicable to DIA, SRP and PRM acquisitions}

@@

\todo{This overlaps with the Datasets section. I would merge all of the below to the experimental data section} All datasets for the four experiments and two simulations, including raw files and analysis results, are available online.

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.6]{images/fig3.png}
\caption{
Goals of PTM characterization, input to statistical analyses, and notation. \todo{Could you update the caption? The figure now has only 2 parts. It doesn't have subscript $i$ etc. On the other hand, it distinguishes the population parameters, and the observed data. Could you mention this here briefly, and describe in more detail in the text?}  (a) Schematic data representation, in a simplified case of two conditions and two \todo{biological?} replicate runs. Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide. The feature intensities the PTM in Condition i are denoted by i*. Features corresponding to unmodified peptides in Condition i is denoted by i. Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines). (b) PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty. (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
\label{fig:data-structure}
\end{figure}
